### 2011-12-19 papers/smectics/smectics.tex
Corrected label for eq. 1
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1500 40876441-3496-0410-a613-a2fca35b2ae2

\documentclass[8.5pt,twoside,twocolumn]{article}
\oddsidemargin -1.2cm
\evensidemargin -1.2cm
\textwidth 18cm
\headheight 1.0in
\topmargin -3.5cm
\textheight 22cm
\usepackage[super,sort&compress,comma]{natbib} 
\usepackage{mhchem}
\usepackage{times,mathptm}
\usepackage{sectsty}
\usepackage{balance} 
\usepackage{graphicx} %eps figures can be used instead
\usepackage{lastpage}
\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
\usepackage{fancyhdr}
\usepackage{bm}
\usepackage{float}

\pagestyle{fancy}

@@

\fancyfoot{}
\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{LF.pdf}}
\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{RF.pdf}}
\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{RF.pdf}}
\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
\fancyhead{}
\renewcommand{\headrulewidth}{1pt} 
\renewcommand{\footrulewidth}{1pt}
\setlength{\arrayrulewidth}{1pt}
\setlength{\columnsep}{6.5mm}
\setlength\bibsep{1pt}

@@

Instances of soft complex fluids abound in our everyday lives: examples include paint (a colloidal suspension), mayonnaise (an emulsion), and engine oil (a polymer solution). Many soft materials, including the amphiphiles used in skin creams, fabric softeners, and some pharmaceutical products, form lamellar mesophases (also known as smectic liquid crystals) whose perfectly ordered state consists of a regular stack of flat layers. As well as these lyotropic liquid crystals arising in amphiphilic solutions, thermotropic smectics commonly arise in systems of rodlike, mesogenic molecules even without solvent [deGennes book]. Such phases are of wide technological importance in devices
such as electronic displays and sensors.

The flow behaviour (rheology) of smectic liquid crystals affects many industrial and manufacturing processes; conversely, rheological measurements provide important probes of their macroscopic material properties.
For a simple fluid the response to steady applied stress is quantified in the linear (Newtonian) regime by a single viscosity. Lamellar phases are anisotropic, and thus in principle have several viscosities as well as an elastic modulus related to layer compression. However, these linear response parameters become somewhat irrelevant when defects are present: unless their density is very low, the defects soon dominate the stress. Moreover, since the defect pattern is strongly dependent on flow rate, the rheology is generically nonlinear. On shearing, the effective viscosity (ratio of shear stress $\sigma_{xy}$  to strain rate $\dot\gamma$) may either increase or decrease, corresponding
to shear thickening or shear thinning behaviour [Larson book]. Also if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) \cite{Olmsted08}.

@@

An intriguing complication in addressing the dynamics of layered systems
is permeation,
where individual molecules in the fluid translate while keeping the
layerwise order intact. In steady state, permeation leads to increased dissipation
and hence to a large viscosity in any geometry where flow necessitates such motion \cite{deGennes}. (A well-known example of permeation flow
is found in cholesterics which are sheared or pushed along the direction
of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction. Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07,Gradzielski03}. (The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear. Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation \cite{Zilman99,Courbin02,Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion \cite{Zipfel99, Leon00}. Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work. Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
phase transitions from a homogeneous state into to ordered phases
like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) play a central role in this transition~\cite{Cates89}. In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
once defects are present).

Despite a wealth of experimental results, and the importance of flow behaviour
to many commercial applications, mechanistic understanding
the rheology of lamellar phases remains limited. This is largely
because of the nonlinear nature of the equations of motion, and the
coupling of the flow to different microstructures. These complications either require
drastic approximations or, as pursued here, use of computer
simulation. However, once defects are present, such simulations are computationally expensive. 
While sheared smectics have been studied extensively via simulations
in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b},
computational constraints have meant almost all these results are limited
to two dimensions. Even in 2D, the number of layers simulated has often not been large enough to meaningfully explore the dynamics of defect textures which can have a controlling influence on rheology. Moreover, for technical reasons a number of these studies have used boundary conditions (BCs) that are not optimal if one wants to predict the behaviour of bulk samples, for example using solid moving walls to drive the fluid flow rather than periodic BCs. This choice does not minimize finite size effects and, when combined with lattice Boltzmann algorithms, severely limits the shear rates attainable in large systems. Moreover, solid walls have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that make it difficult to disentangle bulk and boundary effects.

@@

We describe smectics below via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}. This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations). The free energy is formulated in terms of a scalar order parameter $\phi({\bf r})$. This can be viewed as describing a composition variable in a binary fluid system stabilized by added amphiphiles; one can then speak of alternating layers of fluid and interfaces between these layers. A similar interpretation applies to block copolymers where the composition variable describes the relative local density of the two blocks. However, the same order parameter could equally describe the local number density of the midpoints of a rodlike mesogenic molecule (measured relative to the mean) which can develop a periodic density wave that represents the smectic state.

The governing equations in our model are a convection-diffusion equation
for the scalar order parameter, and the Navier-Stokes equation describing
the hydrodynamics of fluid flow. The stress tensor for the fluid contains not only a Newtonian viscous term but extra contributions arising from local deviations of the order parameter from its state of least free energy. Diffusive relaxation of the order parameter then causes additional dissipation, resulting in an effective viscosity that can be much higher than the intrinsic Newtonian contribution.
Our dynamical equations are solved using a combination of a finite
difference scheme for the convection-diffusion equation, and the
lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation \cite{Marenduzzo07, Henrich10a,  Henrich10b, Henrich11a,Gonnella97}.
We use sliding periodic (Lees-Edwards) boundary conditions for
both the convection-diffusion and Navier-Stokes equations. Note that while these boundary conditions impose a fixed integrated strain rate across the whole system they do not impose a uniform strain rate locally; shear bands can therefore arise if the system so chooses. (As it happens, we do not observe shear bands in the range of shear rates we address here.)

@@

KS: THIS FOLLOWING PARAGRAPH DOESN'T QUITE MAKE SENSE TO ME YET.

Our main finding is that, in two dimensions at least, sheared smectics can show two completely different types of rheological response. At modest rates
of shear, where thermodynamic forces are relatively strong compared with
flow effects,  shear quickly aligns the system into the perfect lamellar phase preferred by thermodynamics,
with only a few defects. The rheology in this regime is quasi-Newtonian, with shear stress linear in the applied shear rate. (There is however a large first normal stress difference.) At higher shear rates, larger thermodynamic forces are generated as layers fail to keep up with the imposed flow. This leads to the nucleation of many new defects, balanced in steady state by annihilation when defects of opposite sign encounter each other. The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters; a comprehensive exploration of parameter space is not possible, but we give a scaling argument which relates $\dot\gamma_c$ to those parameters through an effective capillary number, formed as the ratio of viscous to thermodynamic stress.
Within the defect proliferation regime, the macroscopic rheology is generically nonlinear, and shows shear thinning. The shear stress appears to vary as a sublinear power law, $\sigma_{xy}\propto\dot{\gamma}^m$, with $m\simeq 0.87$ for typical parameter values, although the dependence of $m$ on the parameters precludes interpretation of this as a universal scaling law.
Such a behaviour is qualitatively reminiscent of the one found experimentally in [REFERENCE], where the numerical value of $m$ was however different. In this strongly non-Newtonian regime we also find that the rheology is history dependent.

@@

This paper is organized as follows. In Section \ref{themodel} we introduce the model,
including its equilibrium properties, the full stress tensor and the
dynamical equations. We briefly sketch how we derive structural information
from the order parameter and describe a method for quantitative analysis of the defect density. Section \ref{parametermap} describes the mapping between our simulation parameters and those for physical systems of interest.
We present results of the two-dimensional case in Section \ref{results2D}, discussing
the morphology as a function of flow rate. We also consider average and local shear stress densities and
first normal stress differences, and explore their relation to the defect structure. In addition we report Fourier-space information (structure factors) under flow.
In Section \ref{results3D} we present and compare our results for 3D simulations which, by computational necessity, are less comprehensive than in 2D. 
A summary with conclusions closes the paper in Section \ref{conclusions}.

@@

The equilibrium phase in a system of volume $V$ is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75, Gompper}.
%
\begin{eqnarray}
{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}\nonumber\label{FE-functional}
\end{eqnarray}
%
In minimizing this free energy at equilibrium, we ignore thermal fluctuations (treated in \cite{Brazovskii75,Cates89}), thereby adopting a mean-field approximation as in previous literature \cite{Kendon01,Gonnella97, Gonnella98, Xu03, Xu05, Xu06a, Xu06b}. This should be a good approximation far from the ordering transition at which $\phi$ first becomes nonuniform. The uniform state has $\phi = 0$ and hence ${\cal F}= 0$.

@@

The dynamics of the order parameter is governed by an advection-diffusion equation with a diffusive current of $\phi$ dependent on the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$, as follows:
%
\begin{equation}\label{OPEOM}
\partial_t \phi +{\nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu 
\end{equation}
%
Here $M$ is a mobility coefficient. A wavevector-dependent linearized collective diffusivity $D_k$ can then be defined via $\dot\phi_k = -k^2D_k\phi_k$, whose value at the ordering wavevector obeys $D_q = M(a-a_c)/2$ (this is negative, signifying the instability). In contrast, an estimate of the molecular diffusivity can be made by considering zero wavevector, setting $b = -a$ (see below) and examining the relaxation equation for small deviations around the uniformly ordered state at $\phi = \pm 1$. This gives $\dot\phi = 2bM\nabla^2\phi$ so that the molecular diffusivity is of order $\tilde D = 2Mb$.

Treating the fluid as incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation
%
\begin{equation}\label{NSE}
\rho(\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha) = \partial_\beta \sigma_{\alpha \beta}. 
\end{equation}
%
where $\rho$ is the mass density. The stress tensor in this equation reads
%
\begin{equation}
\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-p\delta_{\alpha\beta} - P_{\alpha \beta}. \label{fullstress}
\end{equation}
%
This consists of a Newtonian hydrodynamic part, which is proportional to the background viscosity $\eta$ and the symmetrized velocity gradient, a hydrostatic pressure term, which includes an isotropic stress contribution from the order parameter 
%
\begin{eqnarray}
p&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
\end{eqnarray}
%
and finally another term consisting of the order-parameter contribution to the deviatoric stress
\cite{Yang76,Evans79} [REFERENCE OK?]
%
\begin{equation}
P_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
\end{equation}

In our hybrid LB scheme, the Navier-Stokes equation (\ref{NSE}) is solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
The velocity field is then used in the advection-diffusion equation Eq.\ref{OPEOM}, which is solved by a finite-difference method at the same timestep. We assume periodic boundary conditions along the flow and vorticity direction and Lees-Edwards sliding periodic boundary conditions\cite{leesedwards,Wagner02} along the direction of the velocity gradient.
Following \cite{Wagner02}, we introduce
additional internal Lees-Edwards planes to overcome an upper limit on shear rate that would otherwise be inversely proportional to the system size in the gradient direction. (This restriction arises because fluid velocities must be small compared to the sound speed in the frame of the lattice; the additional planes create a stack of lattice slabs whose velocity is nearly matched to the local fluid velocity.) 
The total number of Lees-Edwards planes was varied according to the size of the simulation so that the separation of two planes was always 16 lattice sites.

Structural information in Fourier space was obtained from the static structure factor $C({\mathbf k})$, defined as
%
\begin{equation}
C({\mathbf k})=\phi({\mathbf k})\phi(-{\mathbf k}),
\label{S}\end{equation}
%
with $\phi({\mathbf k})$ the normalized Fourier transform of the order parameter.
The transform itself was calculated using the FFTW-library \cite{FFTW}.
In principle an ensemble- or time-average should appear in Eq.~\ref{S} so that $C$ becomes a smooth function of wavevector; in practice an instantaneous measurement, plotted as a colour intensity map, was found sufficient.

>>>>

%\documentclass[8.5pt,twoside,twocolumn]{article}
%\oddsidemargin -1.2cm
%\evensidemargin -1.2cm
%\textwidth 18cm
%\headheight 1.0in
%\topmargin -3.5cm
%\textheight 22cm
%\usepackage[super,sort&compress,comma]{natbib} 
%\usepackage{mhchem}
%\usepackage{times,mathptm}
%\usepackage{sectsty}
%\usepackage{balance} 
%\usepackage{graphicx} %eps figures can be used instead
%\usepackage{lastpage}
%\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
%\usepackage{fancyhdr}
%\usepackage{bm}
%\usepackage{float}

%\pagestyle{fancy}
\documentclass[8.5pt,twoside,twocolumn]{article}
\oddsidemargin -1.2cm
\evensidemargin -1.2cm
\textwidth 18cm
\headheight 1.0in
\topmargin -3.5cm
\textheight 22cm
\usepackage[super,sort&compress,comma]{natbib} 
\usepackage{mhchem}
\usepackage{times,mathptm}
\usepackage{sectsty}
\usepackage{balance}

\usepackage{graphicx} %eps figures can be used instead
\usepackage{lastpage}
\usepackage[format=plain,justification=raggedright,singlelinecheck=false,font=small,labelfont=bf,labelsep=space]{caption} 
\usepackage{fancyhdr}
\usepackage{bm}
\pagestyle{fancy}
\usepackage{color}

@@

\fancyfoot{}
%\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{headers/LF.pdf}}
%\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{headers/RF.pdf}}
%\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{headers/RF.pdf}}
\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
\fancyhead{}
\renewcommand{\headrulewidth}{1pt} 
\renewcommand{\footrulewidth}{1pt}

\setlength{\arrayrulewidth}{1pt}
\setlength{\columnsep}{6.5mm}
\setlength\bibsep{1pt}

\if{\thispagestyle{plain}
\fancypagestyle{plain}{
%\fancyhead[L]{\includegraphics[height=8pt]{LH.pdf}}
%\fancyhead[C]{\hspace{-1cm}\includegraphics[height=20pt]{CH.pdf}}
%\fancyhead[R]{\includegraphics[height=10pt]{RH.pdf}\vspace{-0.2cm}}
\renewcommand{\headrulewidth}{1pt}}
\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\renewcommand\footnoterule{\vspace*{1pt}% 
\hrule width 3.4in height 0.4pt \vspace*{5pt}} 
\setcounter{secnumdepth}{5}

\fancyfoot{}
\fancyfoot[LO,RE]{\vspace{-7pt}\includegraphics[height=9pt]{LF.pdf}}
\fancyfoot[CO]{\vspace{-7.2pt}\hspace{12.2cm}\includegraphics{RF.pdf}}
\fancyfoot[CE]{\vspace{-7.5pt}\hspace{-13.5cm}\includegraphics{RF.pdf}}
\fancyfoot[RO]{\footnotesize{\sffamily{1--\pageref{LastPage} ~\textbar  \hspace{2pt}\thepage}}}
\fancyfoot[LE]{\footnotesize{\sffamily{\thepage~\textbar\hspace{3.45cm} 1--\pageref{LastPage}}}}
\fancyhead{}
\renewcommand{\headrulewidth}{1pt} 
\renewcommand{\footrulewidth}{1pt}
\setlength{\arrayrulewidth}{1pt}
\setlength{\columnsep}{6.5mm}
\setlength\bibsep{1pt}
}\fi

@@

Instances of soft complex fluids abound in our everyday lives: examples include paint (a colloidal suspension), mayonnaise (an emulsion), and engine oil (a polymer solution). Many soft materials, including the amphiphiles used in skin creams, fabric softeners, and some pharmaceutical products, form lamellar mesophases (also known as smectic liquid crystals) whose perfectly ordered state consists of a regular stack of flat layers. As well as these lyotropic liquid crystals arising in amphiphilic solutions, thermotropic smectics commonly arise in systems of rodlike, mesogenic molecules even without solvent~\cite{deGennes}. Such phases are of wide technological importance in devices
such as electronic displays and sensors.

The flow behaviour (rheology) of smectic liquid crystals affects many industrial and manufacturing processes; conversely, rheological measurements provide important probes of their macroscopic material properties.
For a simple fluid the response to steady applied stress is quantified in the linear (Newtonian) regime by a single viscosity. Lamellar phases are anisotropic, and thus in principle have several viscosities as well as an elastic modulus related to layer compression. However, these linear response parameters become somewhat irrelevant when defects are present: unless their density is very low, the defects soon dominate the stress. Moreover, since the defect pattern is strongly dependent on flow rate, the rheology is generically nonlinear. On shearing, the effective viscosity (ratio of shear stress $\sigma_{xy}$  to strain rate $\dot\gamma$) may either increase or decrease, corresponding
to shear thickening or shear thinning behaviour~\cite{Larson}. In addition, if a defect texture is present in the system at rest, this may show a viscosity divergence at small $\dot \gamma$ --- i.e., a yield stress.
Finally, in common with other complex fluids, lamellar phases can undergo transitions into heterogeneous states, for instance to form shear bands (layers of unequal $\sigma_{xy}$ or $\dot\gamma$) \cite{Olmsted08}.

@@

An intriguing complication in addressing the dynamics of layered systems
is permeation,
where individual molecules in the fluid translate while keeping the
layerwise order intact. In steady state, permeation leads to increased dissipation
and hence to a large viscosity in any geometry where flow necessitates such motion \cite{deGennes}. (A well-known example of permeation flow
is found in cholesterics which are sheared or pushed along the direction
of the cholesteric helix.) Permeation is often involved in defect motion such as the growth or shrinkage of incomplete layers at the sites of edge dislocations, and may also be implicated in other forms of topological reconstruction. Such reconstruction arises under shear, sometimes to an extreme degree, for instance in the shear-induced transition from an ordered lamellar stack to an onion phase in lyotropic smectics \cite{Diat93,Panizza96,Iwashita07,Gradzielski03}. (The onion phase comprises a space-filling packing of multilayer vesicles; under some conditions similar defect textures can arise without shear \cite{Gomati87,Boltenhagen92,Fournier94,Ramos04}.) The mechanism underlying such global reconstructions under shear of the layer organization remains unclear. Most theoretical predictions have focussed on buckling and other instabilities of perfectly ordered smectics as the initial trigger for onion formation \cite{Zilman99,Courbin02,Guo2002,Soddemann2004, Stewart2009} but the pathway may also strongly involve the physics of defect motion \cite{Zipfel99, Leon00}. Although we do not see onions in the simulations reported here, in two dimensions we do see a transition from a well ordered phase to a defect-rich phase, whose possible link to onion formation deserves to be explored in future work. Another interesting avenue connects with the observation, in block copolymers and some lyotropic surfactant systems, of a shear-induced
phase transitions from a homogeneous state into ordered phases
like the lamellar one~\cite{Cates89,Koppi93,Fredrickson94}. However, theory suggests that thermal noise (causing local fluctuations about the mean state of order) plays a central role in this transition~\cite{Cates89}. In the present work we focus on the simplest case in which such thermal fluctuations are ignored, and the time evolution of the microstructure is deterministic (though certainly chaotic and unpredictable
once defects are present).

Despite a wealth of experimental results, and the importance of flow behaviour
to many commercial applications, mechanistic understanding of
the rheology of lamellar phases remains limited. This is largely
because of the nonlinear nature of the equations of motion, and the
coupling of the flow to different microstructures. These complications either require
drastic approximations or, as pursued here, use of computer
simulation. However, once defects are present, such simulations are computationally expensive. 
While sheared smectics have been studied extensively via simulations
in the literature~\cite{Swift96,Gonnella97,Gonnella98,Xu03,Xu05,Xu06a,Xu06b},
computational constraints have meant almost all these results are limited
to two dimensions. Even in 2D, the number of layers simulated has often not been large enough to meaningfully explore the dynamics of defect textures which can have a controlling influence on rheology. Moreover, for technical reasons a number of these studies have used boundary conditions (BCs) that are not optimal if one wants to predict the behaviour of bulk samples, for example using solid moving walls to drive the fluid flow rather than periodic BCs. This choice does not minimize finite size effects and, when combined with lattice Boltzmann algorithms, severely limits the shear rates attainable in large systems. Moreover, solid walls for the fluid velocity have sometimes been used in conjunction with periodic BCs for the order parameter \cite{Xu06b} -- a somewhat inconsistent combination that makes it difficult to disentangle bulk and boundary effects.

@@

Below, we describe smectics via a widely-used free energy functional \cite{Gompper, Gonnella97, Xu06b, Cencini07}. This provides a robust generic framework for addressing lamellar ordering without being tied to any specific class of materials (although we do become more specific when estimating suitable parameters for the simulations). The free energy is formulated in terms of a scalar order parameter $\phi({\bf r})$. This can be viewed as describing a composition variable in a binary fluid system stabilized by added amphiphiles; one can then speak of alternating layers of fluid, with interfaces between these layers. A similar interpretation applies to block copolymers where the composition variable describes the relative local density of the two blocks. However, the same order parameter could equally describe the local number density of the midpoints of a rodlike mesogenic molecule (measured relative to the mean) which can develop a periodic density wave that represents a thermotropic smectic state.

The governing equations in our model are a convection-diffusion equation
for the scalar order parameter, and the Navier-Stokes equation describing
the hydrodynamics of fluid flow. The stress tensor for the fluid contains not only a Newtonian viscous term but extra contributions arising from local deviations of the order parameter from its state of least free energy. Diffusive relaxation of the order parameter then causes additional dissipation, resulting in an effective viscosity that can be much higher than the intrinsic Newtonian contribution.
Our dynamical equations are solved using a combination of a finite
difference scheme for the convection-diffusion equation, and the
lattice Boltzmann method (LBM)~\cite{Succi} for the Navier-Stokes equation \cite{Marenduzzo07, Henrich10a,  Henrich10b, Henrich11a,Gonnella97}.
We use sliding periodic (Lees-Edwards) boundary conditions for
both the convection-diffusion and the Navier-Stokes equations. Note that while these boundary conditions impose a fixed integrated strain rate across the whole system they do not impose a uniform strain rate locally; shear bands can therefore arise if the system so chooses. (As it happens, we do not observe shear bands in the range of shear rates we address here.)

@@

Our main finding is that, in two dimensions at least, sheared smectics can show two completely different types of rheological response. At modest rates
of shear, where thermodynamic forces are relatively strong compared with
flow effects,  shear quickly aligns the system into the perfect lamellar phase preferred by thermodynamics,
with only a few defects. The rheology in this regime is quasi-Newtonian, with shear stress linear in the applied shear rate. (There is however a large first normal stress difference.) At higher shear rates, larger thermodynamic forces are generated as layers fail to keep up with the imposed flow. This leads to the nucleation of many new defects, balanced in steady state by annihilation when defects of opposite sign encounter each other. The strain rate $\dot\gamma_c$ needed to enter this nonlinear flow regime depends on thermodynamic and kinetic parameters. A comprehensive exploration of parameter space is not possible, but we give a scaling argument which relates $\dot\gamma_c$ to those parameters through an effective capillary number, formed as the ratio of viscous to thermodynamic stress.
Within the defect proliferation regime, the macroscopic rheology is generically nonlinear, and shows shear thinning. The shear stress appears to vary as a sublinear power law, $\sigma_{xy}\propto\dot{\gamma}^m$, with $m\simeq 0.87$ for typical parameter values, although the dependence of $m$ on the parameters precludes interpretation of this as a universal scaling law.
Such a behaviour is qualitatively reminiscent of the one found experimentally in~\cite{Lu08}, where the numerical value of $m$ was however different. In this strongly non-Newtonian regime we also find that the rheology is history dependent.

@@

This paper is organized as follows. In Section \ref{themodel} we introduce the model,
including its equilibrium properties, the full stress tensor and the
dynamical equations. We briefly sketch how we derive structural information
from the order parameter and describe a method for quantitative analysis of the defect density. Section \ref{parametermap} describes the mapping between our simulation parameters and those for physical systems of interest.
We present results for the two-dimensional case in Section \ref{results2D}, discussing
the morphology as a function of flow rate. We also consider average and local shear stress densities and
first normal stress differences, and explore their relation to the defect structure. In addition we report Fourier-space information (structure factors) under flow.
In Section \ref{results3D} we present and compare our results for 3D simulations which, by computational necessity, are less comprehensive than in 2D. 
A summary with conclusions closes the paper in Section \ref{conclusions}.

@@

The equilibrium phase in a system of volume $V$ is found for an order parameter field $\phi({\bm r})\equiv \phi$ that minimises the Landau-Brazovskii free energy functional \cite{Brazovskii75, Gompper}.
%
\begin{eqnarray}
{\cal F}[\phi] &=& \int d^dx f(\phi,\nabla\phi,\nabla^2\phi) = V\bar f
\label{FE-functional}
\\ &=& \int d^dx\left\{\frac{a}{2}\phi^2+\frac{b}{4}\phi^4+\frac{\kappa}{2} ({\nabla}\phi)^2+\frac{c}{2}(\nabla^2\phi)^2\right\}.
\nonumber
\end{eqnarray}
%
In minimizing this free energy at equilibrium, we ignore thermal fluctuations (treated in \cite{Brazovskii75,Cates89}), thereby adopting a mean-field approximation as in previous literature \cite{Kendon01,Gonnella97, Gonnella98, Xu03, Xu05, Xu06a, Xu06b}. This should be a good approximation far from the ordering transition at which $\phi$ first becomes nonuniform. The uniform state has $\phi = 0$ and hence ${\cal F}= 0$.

@@

The dynamics of the order parameter is governed by an advection-diffusion equation with a diffusive current of $\phi$ dependent on the chemical potential $\mu$, the functional derivative of ${\cal F}$ with respect to $\phi$, as follows:
%
\begin{equation}\label{OPEOM}
\partial_t \phi +{\nabla}({\bm u}\phi) = M \nabla^2 \frac{\delta \cal F}{\delta \phi}=M \nabla^2 \mu.
\end{equation}
%
Here $M$ is a mobility coefficient. A wavevector-dependent linearized collective diffusivity $D_k$ can then be defined via $\dot\phi_k = -k^2D_k\phi_k$, whose value at the ordering wavevector obeys $D_q = M(a-a_c)/2$ (this is negative, signifying the instability). In contrast, an estimate of the molecular diffusivity can be made by considering zero wavevector, setting $b = -a$ (see below) and examining the relaxation equation for small deviations around the uniformly ordered state at $\phi = \pm 1$. This gives $\dot\phi = 2bM\nabla^2\phi$ so that the molecular diffusivity is of order $\tilde D = 2Mb$.

Treating the fluid as incompressible, the evolution of the velocity ${\bm u}$ obeys the Navier-Stokes equation
%
\begin{equation}\label{NSE}
\rho(\partial_t u_\alpha + (u_\beta \nabla_\beta) u_\alpha) = \partial_\beta \sigma_{\alpha \beta},
\end{equation}
%
where $\rho$ is the mass density. The stress tensor in this equation reads
%
\begin{equation}
\sigma_{\alpha \beta}=\eta (\partial_\alpha u_\beta + \partial_\beta u_\alpha)-p\delta_{\alpha\beta} - P_{\alpha \beta}. \label{fullstress}
\end{equation}
%
This consists of a Newtonian hydrodynamic part, which is proportional to the background viscosity $\eta$ and the symmetrized velocity gradient, a hydrostatic pressure term, which includes an isotropic stress contribution from the order parameter 
%
\begin{eqnarray}
p_{\phi}&=& \frac{a}{2}\phi^2+\frac{3b}{4}\phi^4-\kappa\left(\phi(\nabla^2\phi)+\frac{1}{2}({\nabla}\phi)^2\right)\nonumber\\
& &\quad+c\left(\phi(\nabla^2)^2\phi+\frac{1}{2}(\nabla^2\phi)^2+\partial_{\gamma}\phi\partial_\gamma(\nabla^2\phi)\right)
\end{eqnarray}
%
and finally another term consisting of the order-parameter contribution to the deviatoric pressure~\cite{Swift96,Yang76,Evans79} 
%
\begin{equation}
P_{\alpha \beta}=\kappa\partial_\alpha\phi\partial_\beta\phi -c \left(\partial_\alpha\phi\partial_\beta(\nabla^2\phi)+\partial_\beta\phi\partial_\alpha(\nabla^2\phi)\right).
\end{equation}

In our hybrid LB scheme, the Navier-Stokes equation (\ref{NSE}) is solved at every timestep by means of a multi-relaxation time lattice-Boltzmann solver \cite{dHumieres02,Adhikari05} using the D3Q19 model.
The velocity field is then used in the advection-diffusion equation Eq.\ref{OPEOM}, which is solved by a finite-difference method at the same timestep. We assume periodic boundary conditions along the flow and vorticity directions and Lees-Edwards sliding periodic boundary conditions\cite{leesedwards,Wagner02} along the direction of the velocity gradient.
Following \cite{Wagner02}, we introduce
additional internal Lees-Edwards planes to overcome an upper limit on shear rate that would otherwise be inversely proportional to the system size in the gradient direction. (This restriction arises because fluid velocities must be small compared to the sound speed in the frame of the lattice; the additional planes create a stack of lattice slabs whose velocity is nearly matched to the local fluid velocity.) 
The total number of Lees-Edwards planes was varied according to the size of the simulation so that the separation of two planes was always 16 lattice sites.
We relegate further comments on the implementation of the Lees-Edwards boundary
conditions to an Appendix.

Structural information in Fourier space was obtained from the static structure factor $C({\mathbf k})$, defined as
%
\begin{equation}
C({\mathbf k})=\phi({\mathbf k})\phi(-{\mathbf k}),
\label{S}\end{equation}
%
with $\phi({\mathbf k})$ the normalized Fourier transform of the order parameter.
The transform itself was calculated using the FFTW-library \cite{FFTW}.
In principle an ensemble-average or time-average should appear in Eq.~\ref{S} so that $C$ becomes a smooth function of wavevector; in practice an instantaneous measurement, plotted as a colour intensity map, was found sufficient.

For the analysis of the relation between rheological quantities and the morphology of the system we need to quantify the number of defects.
For this purpose we defined an alignment function $X({\mathbf r})$, which measures the average orientation of the order parameter gradients, as follows
%
\begin{equation}\label{alignment}
X({\mathbf r})=\frac{1}{V'({\mathbf r})} \int_{V'({\mathbf r})} d{\mathbf r'} \left(\frac{\nabla \phi ({\mathbf r}) \cdot \nabla \phi ({\mathbf r'})}{|\nabla\phi ({\mathbf r}) ||\nabla\phi ({\mathbf r'}) |}\right)^2. 
\end{equation}
%
Here $V'$ is a small volume that contains $\mathbf r$; in practice $V'$ can be chosen to contain $\mathbf r$ and its nearest neighbours only.
